Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-06-05 main.tex
Move figure, spelling, results text.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a suprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by  relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
that applying a motor acutated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
they are likely to fall.

In the early years of developments in automatic control, \citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize the a
bicycle. Much later, automatic roll stabilization of single track vehicles
began in earnest after predictive motorcycle models were developed and refined
in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
work and seems to be the first to attempt to robotically stabilize a small
motorbike with a controlled inverted pendulum that mimicked rider lean, but he
was not successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} sucessfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, somethign not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance assist but contrarily researchers with a
similar vehicle recently showed rider disatisfction with the
stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stablity. The bicycle model is ideal for showing the effect of a roll
motion feedback driven steer motor on the dynamics. The linear model can be
described by the fourth order state space equations:

\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilbirum forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic mulitbody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

@@

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0
\quad k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts
for \(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to
\(v\), the speed range where the bicycle is stable can be maximized within any
physical acutator magnitude and bandwidth limits.  \citet{Schwab2008}
elaborates on some of the possibilities in scheduling the gains for such a
controller on this model.

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle Grenoble C7\todo{Double check
this is the right model}, Dieren, The Netherlands) with a custom motor in the
steering assembly capable of applying up to 7~\si{\newton\meter} of torque
between the head tube and steer tube, Figure~\ref{fig:balance-assist-bicycle}.
A custom motor controller converts the commanded torque to applied torque. We
measure the speed of the rear wheel with an encoder (Magura XXX) and measure
the body fixed roll rate of the bicycle with a MEMs rate gyroscope (Company,
Location). The balance assist control algorithm is implemented on a
microprocessor (Teensy, Company, Location) and data from all sensors is logged
with a CAN bus (Company, Location) at approximately X Hz.
%
\begin{figure}
  \centering
  \includegraphics[width=100mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with steer motor and data
    aquisition and control electronics mounted in the rear rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino XXX, Italy). We measure the force applied by each motor
at the handlebar via four inline load cells, each with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangment
is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
to that described in the thesis include simplifying the controller with a
inexpensive microcontroller and the use of a simpler safety harness.
Figure~\ref{fig:perturbation_0} shows an example perturbation.
%
% NOTE : I get a compile error with the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

@@

\subsection{Protocol}
%
The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}) with the controller gain at $k=8$ and the second
group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occured X
weeks before the 6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
location of their front wheel on the center line of the treadmill. We define a
``fall'' on the treadmill by two criteria: the rider removes their foot from
the pedal and places it on the ground or the bicycle wheel exceeds the width of
the treadmill. We then applied perturbations in random directions (clockwise or
counter clockwise) increasing the magnitude by X~\si{\newton\meter} each time.
Figure~\ref{fig:perturbation-sequence} shows an example reult of a
perturbation.  We log the magnitude that causes the first fall to characterize
that subject's nominal fall threshold.  Following this we apply 20
perturbations of random magnitudes to the cyclist while they ride at a constant
speed and record which perturbations cause a fall. We let the cyclist rest and
then perform another 20 perturbations. We randomize whether the balance assist
system is on or off during the first or second set of perturbations.
%
\begin{figure}
  \centering
  \subcaptionbox{Start}{\includegraphics[width=40mm]{figures/start.png}}
  \subcaptionbox{During}{\includegraphics[width=40mm]{figures/during.png}}
  \subcaptionbox{After}{\includegraphics[width=40mm]{figures/after.png}}
  \subcaptionbox{Recovery}{\includegraphics[width=40mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

\begin{align}
  F_r = & F_{rf} - F_{rr} \\
  F_l = & F_{lf} - F_{lr} \\
  L = & \int_{0\si{\second}}^{0.3\si{\second}} l/2(F_r + F_l) dt
  \textrm{.}
\end{align}

@@

\subsection{Statistics}
%
We test the hypothesis that the balance assistance controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary outcome variable \(f\) that is dependent on several possible
explanatory independent variables including the binary balance assistance (on
or off).

\begin{table}
  \centering
  \caption{Statisitcal model variables.}
  \begin{tabular}{lll}
    \toprule
    Variable & Units & Description \\
    \midrule
    \(L\) & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(c\) & integer & order number of perturbation \\
    \(k\) & TODO & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (1.7 or 2.8) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
    \bottomrule
  \end{tabular}
\end{table}

We evaluate this hypothesis using a multivariate logistic regression model in
the form

\begin{align}
  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij}) \textrm{.}
\end{align}

Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) of
participant \(i\) which follows a Bernoulli distribution given the probability
\(p_{ij}\) that a fall occurs. The log-odds of the probabiliy is then a linear
function of our independent variables with \(\beta\) as the intercept and
\(\alpha_k\) as the linear coefficients.

\begin{align}
  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) = \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
\end{align}

@@

We scale all independent variables such they they have a mean of zero and a
standard deviation of one. We use cluster-mean centering, as recommended by
\cite{Enders2007}, with the clusters being an individual subject because we are
interested in the association between the state of the balance-assist system
and the outcome of the perturbation.

Cluster-mean centering shows there to be no variation between participants. We
expected there to be a variation between participants in how well they are able
to resist the perturbation. However, this was not true. This allowed us to
stick with a simple single-level logistic regression model, instead of a
multilevel model. This left use with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independen tvariables.. We also include interaction effects between the balance
assist state and the other four variables.

We divide the analysis into two separate model evalautions, one for the
6~\si{\kilo\meter\per\hour} \(k=10\) trials and one for the
10~\si{\kilo\meter\per\hour} \(k=8\) trials and we evaluate our hypothesis for
each set of data.

@@

\begin{figure}
  \centering
  \includegraphics{example-image-a}
  \caption{Box plot or distrubtion plot showing graphical view of stats
  tables.}
\end{figure}

@@

Angular impulse magnitude has the largest significant effect for predicting
fall probability. An increase in angular impulse increases the fall probability
both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds
is approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a
more important predictor at higher speeds compared to lower speeds. The reason
for this likely has to do with the width of the treadmill. As a bicycle travels
at higher speeds, the same perturbation causes larger lateral deviations. At
10~\si{\kph} almost all falls were due to the bicycle exiting the maximum width
of the treadmill. If the same experiment was performed on an infinite plane,
the riders may have recovered from the perturbation. At 6~\si{\kph} the riders
could often recover in the alloted treadmill width. Our results are very much
dependent on the two modes of falling with use: exit the treadmill width or
foot is placed on the belt. Cycle paths are a similar width as the treadmill,
so ride's are often limited in width when recoverying from a fall.

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by  relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
that applying a motor actuated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
%
\begin{figure}
  \centering
  \includegraphics[width=70mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with steer motor and data
    acquisition and control electronics mounted in the rear rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

In the early years of developments in automatic control, \citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize a
bicycle. Much later, automatic roll stabilization of single track vehicles
began in earnest after predictive motorcycle models were developed and refined
in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
work and seems to be the first to attempt to robotically stabilize a small
motorbike with a controlled inverted pendulum that mimicked rider lean, but he
was not successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, something not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance assist but contrarily researchers with a
similar vehicle recently showed rider dissatisfaction with the
stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stability. The bicycle model is ideal for showing the effect of a roll
motion feedback driven steer motor on the dynamics. The linear model can be
described by the fourth order state space equations:

\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilibrium forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic multibody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

@@

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0
\quad k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts
for \(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to
\(v\), the speed range where the bicycle is stable can be maximized within any
physical actuator magnitude and bandwidth limits.  \citet{Schwab2008}
elaborates on some of the possibilities in scheduling the gains for such a
controller on this model.

@@

\todo{Talk about the types of real life perturbations an how our perturbation
mimics it.}

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle Grenoble C7\todo{Double check
this is the right model}, Dieren, The Netherlands) with a custom motor in the
steering assembly capable of applying up to 7~\si{\newton\meter} of torque
between the head tube and steer tube, Figure~\ref{fig:balance-assist-bicycle}.
A custom motor controller converts the commanded torque to applied torque. We
measure the speed of the rear wheel with an encoder (Magura XXX) and measure
the body fixed roll rate of the bicycle with a MEMs rate gyroscope (Company,
Location). The balance assist control algorithm is implemented on a
microprocessor (Teensy, Company, Location) and data from all sensors is logged
with a CAN bus (Company, Location) at approximately X Hz.

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino XXX, Italy). We measure the force applied by each motor
at the handlebar via four inline load cells, each with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangement
is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
to that described in the thesis include simplifying the controller with a
inexpensive microcontroller and the use of a simpler safety harness.
Figure~\ref{fig:perturbation_0} shows an example perturbation.
%
% NOTE : I get a compile error with the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

@@

\subsection{Protocol}
%
The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}) with the controller gain at $k=8$ and the second
group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred X
weeks before the 6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
location of their front wheel on the center line of the treadmill. We define a
``fall'' on the treadmill by two criteria: the rider removes their foot from
the pedal and places it on the ground or the bicycle wheel exceeds the width of
the treadmill. We then applied perturbations in random directions (clockwise or
counter clockwise) increasing the magnitude by X~\si{\newton\meter} each time.
Figure~\ref{fig:perturbation-sequence} shows an example result of a
perturbation.  We log the magnitude that causes the first fall to characterize
that subject's nominal fall threshold.  Following this we apply 20
perturbations of random magnitudes to the cyclist while they ride at a constant
speed and record which perturbations cause a fall. We let the cyclist rest and
then perform another 20 perturbations. We randomize whether the balance assist
system is on or off during the first or second set of perturbations.
%
\begin{figure}
  \centering
  \subcaptionbox{Start}{\includegraphics[width=40mm]{figures/start.png}}
  \subcaptionbox{During}{\includegraphics[width=40mm]{figures/during.png}}
  \subcaptionbox{After}{\includegraphics[width=40mm]{figures/after.png}}
  \subcaptionbox{Recovery}{\includegraphics[width=40mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \textrm{.}
\end{align}

@@

\subsection{Statistics}
%
We test the hypothesis that the balance assistance controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary fall outcome variable \(f\) that is dependent on several possible
explanatory independent variables, one of which is the binary balance
assistance (on or off). See Table~\ref{tab:stat-model-variables} for all of the
variables.
%
\begin{table}
  \centering
  \caption{Statistical model variables.}
  \label{tab:stat-model-variables}
  \begin{tabular}{lll}
    \toprule
    Variable & Units & Description \\
    \midrule
    \(L\) & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(c\) & integer & order number of perturbation \\
    \(k\) & TODO & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (1.7 or 2.8) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
    \bottomrule
  \end{tabular}
\end{table}

We evaluate this hypothesis using a multivariate logistic regression model that
takes the general form
%
\begin{align}
  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij}) \textrm{.}
\end{align}
%
Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
participant \(i\) which follows a Bernoulli distribution given the probability
\(p_{ij}\) that a fall occurs. The log-odds of the probability is then a linear
function of our independent variables with \(\beta\) as the intercept and
\(\alpha_k\) as the linear coefficients to the independent variables
\(x_{ij}^k\), i.e. all variables in Table~\ref{tab:statmodel-variables} except
\(f\).
%
\begin{align}
  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) = \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
\end{align}

@@

Before fitting the model, we scale each \(x_{ij}^k\) such they they have a mean
of zero and a standard deviation of one by cluster-mean centering, as
recommended by \cite{Enders2007}, with clusters being an individual subject.
The clusters are all data from an individual subject because we are only
interested in the association between the state of the balance-assist system
and the outcome of the perturbation. We expected there to be a variation
between participants in how well they are able to resist the perturbation.
However, this was not true.  Cluster-mean centering shows there to be no
variation between participants. This allowed us to stick with a simple
single-level logistic regression model, instead of a multilevel model. This
left us with angular impulse, perturbation order, balance assist state, and
roll \& steer angles at the time of perturbation as independent variables. We
also include interaction effects between the balance assist state and the other
four variables.

We divide the analysis into two separate model evaluations, one for the
6~\si{\kilo\meter\per\hour} \(k=10\) trials and one for the
10~\si{\kilo\meter\per\hour} \(k=8\) trials and we evaluate our hypothesis for
each set of data.

@@

\begin{figure}
  \centering
  \includegraphics{example-image-a}
  \caption{Box plot or distribution plot showing graphical view of stats
  tables.}
\end{figure}

@@

Angular impulse magnitude has the largest significant effect for predicting
fall probability. An increase in angular impulse increases the fall probability
both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds
is approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a
more important predictor at higher speeds compared to lower speeds. The reason
for this likely has to do with the width of the treadmill. As a bicycle travels
at higher speeds, the same perturbation causes larger lateral deviations. At
10~\si{\kph} almost all falls were due to the bicycle exiting the maximum width
of the treadmill. If the same experiment was performed on an infinite plane,
the riders may have recovered from the perturbation. At 6~\si{\kph} the riders
could often recover in the allotted treadmill width. Our results are very much
dependent on the two modes of falling with use: exit the treadmill width or
foot is placed on the belt. Cycle paths are a similar width as the treadmill,
so ride's are often limited in width when recovering from a fall.
